\artus{} -- Framework for Event-based Data Analysis in High Energy Physics

### 2015-10-27 Core/doc/latex/paper.tex
Paper: Move table inside chapter
I think putting the table directly before the reference in the text
is acceptable, better than in the conclusion.
Paper: First figure caption

The data which are relevant for a specific analysis are selected in a step usually referred to as \textit{skimming}.
The output of the skimming process is an analysis-dependent file, usually in an n-tuple format, and small enough to be stored on local file servers, often owned and operated by a single institute.
The size of these files ranges from several \SI{10}{GB} to few~TB.

@@

\begin{figure}[t]
\centering \input{figures/analysis_steps}
\caption[Characteristics of an event-based analysis in high energy physics]{Characteristic of an event-based analysis in high energy physics, numbers taken from \cite{joram_phd}.
Data size is reduced by six orders of magnitude, turnaround time by a factor of 4.
I/O intensive operations are done in highly parallel in the WLCG.
The data reduction is achieved by performing the data-intensive operations only rarely and applying sensible event selection steps as early as possible.}
\label{figure_analysis_steps}
\end{figure}

@@

\begin{description}
\item[Performance]
The main performance indicator is the \emph{turnaround time} which is the time needed to run the whole analysis in case new requirements arise.
Since the runtime is often limited by the transfer speed of the storage servers, the software should load each event only once and perform all necessary processing steps.
Still, since analysis requirements sometimes change within a few hours, a high level of flexibility is needed.
This explains the need for a modular structure, where only those parts of the code are executed that are really necessary for the analysis.

\item[Maintainability]
Source code maintenance, i.e.\ to be up-to-date with the latest requirements while not losing backwards compatibility, is a challenge for each software project.
The first important step is to always have an updated recipe that works out of the box on each machine offering CVMFS.
To make the entry level as low as possible, beginners can start with an example analysis.
In \artus, everything is steered via JSON configuration files.
The modules then implement what has to be done while the data is stored in desired data format.

@@

\artus respects the separation of code and configuration.
Settings select what has to be done and with which parameters.
The code does perform these steps, ideally without the user being fully aware of implementation details.
The single source of truth for an analysis is therefore a JSON file that completely defines the analysis.
It contains information on the input data and settings for the pipelines and processors described in the following.
Each of the modules described in the following has access to the settings read in from the configuration file.
The \artus package comes with sophisticated tools written in python, that help in generating this JSON file.

@@

The events are then sequentially processed by processors organized in pipelines.
First, a so-called global pipeline is run.
Then, the execution is split up into multiple so-called local pipelines, that are also processed sequentially.
The composition of a pipeline is illustrated in Figure~\ref{figure_artus_pipeline}.

@@

\section{Building an End-user Analysis}
\label{section_artus_analysis}
In order to create a custom analysis, the analysis programmer can build his own code on the \artus classes by deriving from them.
He must provide an implementation of the settings, the event and the product class.
The settings class defines the tags that are read from the configuration file and the corresponding types of these values.
The \emph{event} class defines the event content which is read from the input and the product class contains members for new quantities calculated during the analysis stage.
These three classes need to be derived from the base classes provided by the core of the \artus framework.

@@

@@

%time HiggsToTauTauAnalysis.py @HiggsAnalysis/KITHiggsToTauTau/data/ArtusWrapperConfigs/Sync13TeV.cfg -i HiggsAnalysis/KITHiggsToTauTau/data/Samples/NAF_sample_DYJetsToLLM50_RunIISpring15DR74_Asympt25ns_13TeV_MINIAOD_amcatnloFXFX-pythia8_recent.txt -f 10
\begin{table}[h!]
\centering
\begin{tabular}{lrr}
\toprule
\textbf{Pipelines}                                & \textbf{Runtime / \si{min}} & \textbf{Event Rate / \si{s^{-1}} } \\ \midrule
%0                                                & 5:53                             & 1296 \\
 0                                                & 5:12                             & 1470 \\ \midrule
%\em                                              & 6:27                             & 1181 \\
 1 ($\Pe\Pe$)                                     & 6:31                             & 1174 \\
%1 ($\Pe\Pgt$)                                    & 7:03                             & 1084 \\
 1 ($\Pe\Pgt$)                                    & 7:26                             & 1028 \\
%1 ($\Pgm\Pgt$)                                   & 7:14                             & 1054 \\
 1 ($\Pgm\Pgt$)                                   & 7:14                             & 1052 \\
%1 ($\Pgt\Pgt$)                                   & 6:14                             & 1224 \\
 1 ($\Pgt\Pgt$)                                   & 6:07                             & 1247 \\ \midrule
%2 ($\Pgm\Pgt$, $\Pe\Pgt$)                        & 8:33                             & 893 \\
 2 ($\Pgm\Pgt$, $\Pe\Pgt$)                        & 8:32                             & 896 \\
%3 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$)             & 9:59                             & 760 \\
 3 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$)             & 9:54                             & 767 \\
%4 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$, $\Pgt\Pgt$) & 10:46                            & 712 \\
 4 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$, $\Pgt\Pgt$) & 10:01                            & 758 \\
\bottomrule
\end{tabular}
\caption[Measurement of the runtime of the baseline $\PH\to\Pgt\Pgt$ analysis.]{Measurement of the runtime of the baseline $\PH\to\Pgt\Pgt$ analysis.
The analysis is performed on 10 files with a total size of 6.1~GB containing 452229 simulated $\PZ+\text{jet}$ events.
The measurement is performed with different numbers of pipelines, where each pipeline implements the analysis of one sub-channel of the analysis.
The values are compared with the execution without any pipelines and without global processors.
It is clearly visible, that the time needed for reading the inputs amounts to a large part of the overall runtime and that the simultaneous execution of multiple pipelines can reduce the runtime significantly compared to the case where different configuration are performed sequentially including separate IO operations.}
\label{table_artus_runtime_comparison}
\end{table}

@@

Especially new students joining the analysis groups can contribute very quickly.
Once they have understood general concept and structure of an \artus analysis, they can contribute by writing their own producers or filters while the full complexity of the event-processing is taken care of by the \artus framework.
In a similar manner, using the \artus framework, it is now much less work to set up a new analysis because a rich library of processors, filters and consumers exists and can be reused for new analysis.

>>>>

The data which are relevant for a specific analysis are selected in a step usually referred to as \textit{skimming}.
The output of the skimming process is an analysis-dependent file\todo[inline]{better "format" instead of "file"? (needs restructuring of the sentence a bit}, usually in an n-tuple format, and small enough to be stored on local file servers, often owned and operated by a single institute.
The size of these files ranges from several \SI{10}{GB} to few~TB.

@@

\begin{figure}[t]
\centering \input{figures/analysis_steps}
\caption[Characteristics of an event-based analysis in high energy physics]{Characteristic of an event-based analysis in high energy physics, numbers taken from \cite{joram_phd}.
Data size is reduced by six orders of magnitude, the turnaround time by a factor of 10000.
I/O intensive operations are done highly parallelized in the WLCG.
The data reduction is achieved by performing the data-intensive operations only rarely and applying analysis specific event selection steps as early as possible.}
\label{figure_analysis_steps}
\end{figure}

@@

\begin{description}
\item[Performance]
The main performance indicator is the \emph{turnaround time} which is the time needed to run the whole\todo[inline]{"entire" always better than "whole"?} analysis in case new requirements arise.
Since the runtime is often limited by the transfer speed of the storage servers, the software should load each event only once and perform all necessary processing steps.
Still, since analysis requirements sometimes change within a few hours, a high level of flexibility is needed.
This explains the need for a modular structure, where only those parts of the code are executed that are really necessary for the analysis.

\item[Maintainability]
Source code maintenance, i.e.\ to be up-to-date with the latest requirements while not losing backwards compatibility, is a challenge for each software project.
\todo[inline]{Didn't we agree to remove the following sentence?} The first important step is to always have an updated recipe that works out of the box on each machine offering CVMFS.
To make the entry level as low as possible, beginners can start with an example analysis.
In \artus, everything is steered via JSON configuration files.
The modules then implement what has to be done while the data is stored in desired data format.

@@

\artus respects the separation of code and configuration.
Settings select\todo[inline]{better "specify" instead of "select"?} what has to be done and with which parameters.
The code does perform these steps, ideally without the user being fully aware of implementation details.
The single source of truth for an analysis is therefore a JSON file that completely defines the analysis.
It contains information on the input data and settings for the pipelines and processors described in the following.
Each of the modules described in the following has access to the settings read in from the configuration file.
The \artus package comes with sophisticated tools written in python, that help in generating this JSON file.

@@

The events are then sequentially processed by processors organized in pipelines.
First, a so-called global pipeline is run.
Then,\todo[inline]{No comma here?} the execution is split up into multiple so-called local pipelines, that are also processed sequentially.
The composition of a pipeline is illustrated in Figure~\ref{figure_artus_pipeline}.

@@

\section{Building an End-user Analysis}
\label{section_artus_analysis}
In order to create a custom analysis, the analysis programmer can build his own code on the \artus classes by deriving from them.
He must provide an implementation of the settings, the event and the product class.
The settings class defines the tags that are read from the configuration file and the corresponding types of these values.\todo[inline]{better "and the types of the corresponding values"?}
The \emph{event} class defines the event content which is read from the input and the product class contains members for new quantities calculated during the analysis stage.
These three classes need to be derived from the base classes provided by the core of the \artus framework.

@@

%time HiggsToTauTauAnalysis.py @HiggsAnalysis/KITHiggsToTauTau/data/ArtusWrapperConfigs/Sync13TeV.cfg -i HiggsAnalysis/KITHiggsToTauTau/data/Samples/NAF_sample_DYJetsToLLM50_RunIISpring15DR74_Asympt25ns_13TeV_MINIAOD_amcatnloFXFX-pythia8_recent.txt -f 10
\begin{table}[!ht]
\centering
\begin{tabular}{lrr}
\toprule
\textbf{Pipelines}                                & \textbf{Runtime / \si{min}} & \textbf{Event Rate / \si{s^{-1}} } \\ \midrule
%0                                                & 5:53                             & 1296 \\
 0                                                & 5:12                             & 1470 \\ \midrule
%\em                                              & 6:27                             & 1181 \\
 1 ($\Pe\Pe$)                                     & 6:31                             & 1174 \\
%1 ($\Pe\Pgt$)                                    & 7:03                             & 1084 \\
 1 ($\Pe\Pgt$)                                    & 7:26                             & 1028 \\
%1 ($\Pgm\Pgt$)                                   & 7:14                             & 1054 \\
 1 ($\Pgm\Pgt$)                                   & 7:14                             & 1052 \\
%1 ($\Pgt\Pgt$)                                   & 6:14                             & 1224 \\
 1 ($\Pgt\Pgt$)                                   & 6:07                             & 1247 \\ \midrule
%2 ($\Pgm\Pgt$, $\Pe\Pgt$)                        & 8:33                             & 893 \\
 2 ($\Pgm\Pgt$, $\Pe\Pgt$)                        & 8:32                             & 896 \\
%3 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$)             & 9:59                             & 760 \\
 3 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$)             & 9:54                             & 767 \\
%4 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$, $\Pgt\Pgt$) & 10:46                            & 712 \\
 4 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$, $\Pgt\Pgt$) & 10:01                            & 758 \\
\bottomrule
\end{tabular}
\caption[Measurement of the runtime of the baseline $\PH\to\Pgt\Pgt$ analysis.]{Measurement of the runtime of the baseline $\PH\to\Pgt\Pgt$ analysis.
The analysis is performed on 10 files with a total size of 6.1~GB containing 452229 simulated $\PZ+\text{jet}$ events.
The measurement is performed with different numbers of pipelines, where each pipeline implements the analysis of one sub-channel of the analysis.
The values are compared with the execution without any pipelines and without global processors.
It is clearly visible, that the time needed for reading the inputs amounts to a large part of the overall runtime and that the simultaneous execution of multiple pipelines can reduce the runtime significantly compared to the case where different configuration are performed sequentially including separate IO operations.}
\label{table_artus_runtime_comparison}
\end{table}

@@

@@

Especially new students\todo[inline]{Das klingt ja schon fast, als st\"unden am Ende "echte" Menschen hinter den besprochenen Dingen. F\"ur Paper eher ungew\"ohnlich?} joining the analysis groups can contribute very quickly.
Once they have understood general concept and structure of an \artus analysis, they can contribute by writing their own producers or filters while the full complexity of the event-processing is taken care of by the \artus framework.
In a similar manner, using the \artus framework, it is now much less work to set up a new analysis because a rich library of processors, filters and consumers exists and can be reused for new analysis.
